Algorithms for Estimating Trends in Global Temperature Volatility

### 2018-09-05 aaai-2018/khodadadi-mcdonald-aaai19.tex
proof

\pdfinfo{
  /Title (Algorithms for Estimating Trends in Global Temperature Volatility)
  % /Author (Khodadadi and McDonald)
  /Author (John Doe, Jane Doe)}

@@

\begin{abstract}
Trends in terrestrial temperature variability are perhaps more relavant for species
viability than trends in mean
temperature. In this paper, we develop methodology for estimating such
trends using multi-resolution climate data from polar orbiting weather
satellites. We derive two novel algorithms for computation that are
tailored for dense, gridded observations over both 
space and time. We evaluate our methods with a simulation constructed
to mimic these data's 
features and on a large, publicly available, global
temperature dataset with the eventual goal of tracking trends in
cloud reflectance temperature variability.
\end{abstract}

@@

Trends in terrestrial temperature variability are perhaps more relavant for species
viability than trends in mean
temperature~\cite{huntingford_no_2013},
because an increase in
temperature variability will increase the probability of extreme hot
outliers~\cite{VasseurDeLong2014}. Recent climate literature
suggests that it is more difficult for society to adapt to these
extremes than to the gradual increase in the mean temperature
\cite{hansen_perception_2012,huntingford_no_2013}. Furthermore, the willingness of popular media to
emphasize the prevalence extreme cold events coupled with a
fundamental misunderstanding of the relationship between climate (the global
distribution of weather over the long run) and weather (observed
short-term, localized behavior) leads to public misunderstanding of climate
change. In fact, a point of active debate is the extent to which the observed
increased frequency of extreme cold events in the 
northern hemisphere can be attributed to increases in temperature
variance  rather than to increases in
mean climate~\cite{Screen2014,FischerBeyerle2013,TrenberthZhang2014}.

@@

\citet{KimKoh2009} proposed $\ell_1$-trend filtering as a
method for estimating a smooth, time-varying trend. It is formulated
as the optimization problem 
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that
% it consists of a smooth trend and a stochastic component and the
% goal is to estimate the trend. The estimated trend is desired to be
% smooth, but at the same time, we seek small estimated residuals
% (stochastic component). These two objectives cannot be achieved
% simultaneously and so a trade-off between them should be made. In
% $\ell_1$-trend filtering, this trade-off is formulated as the
% following convex optimization problem: 
$$
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda
\sum_{t=1}^{T-2} \left|\beta_t-2\beta_{t+1}+\beta_{t+2}\right| 
$$
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \norm{ y-\beta }_2^2+\lambda \norm{ D_t \beta}_1
\label{eq:l1tf}
\end{equation}
 where $y=\{y_t\}_{t=1}^T$ is an observed time-series, $\beta\in \mathbb{R}^T$ is the smooth trend,
 $D_t$ is a $(T-2)\times T$ matrix, and $\lambda$ is a tuning parameter
 which balances fidelity to the data (small errors in the first term)
 with a desire for smoothness.  
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
% In Appendix A, we specify the form of the matrix $D$ which results in piecewise
% linear estimated $\beta$. 
\citet{KimKoh2009} proposed a specialized primal-dual
interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a
statistical perspective, \eqref{eq:l1tf} can be viewed as a constrained maximum
likelihood problem with independent observations from a normal
distribution with common variance, $y_t \sim \mbox{N}(\beta_t,
\sigma^2)$, subject to a piecewise linear constraint on
$\beta$. Alternatively, solutions to \eqref{eq:l1tf} are %equivalent to
maximum a posteriori Bayesian estimators based on Gaussian likelihood
with a special Laplace prior distribution on $\beta$. Note that the
structure of the estimator is determined by the penalty function
$\lambda\norm{D_t\beta}_1$ rather than any parametric trend
assumptions---autoregressive, moving average, sinusoidal seasonal
component, etc. The resulting trend is therefore essentially
nonparametric in the same way that splines are nonparametric. In fact,
using squared $\ell_2$ norm as the penalty instead of $\ell_1$ results exactly in
regression splines.

@@

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend filtering on
graphs~\cite{WangSharpnack2016} which penalizes the difference between
the estimated values of the signal on the connected nodes (though the
likelihood is different). As before,
we can write \eqref{eq:l1tf_var_st} in matrix form where $h$ is a
vector of length
$TS$ and $D_t$ is replaced by $D \in
\mathbb{R}^{(N_t+N_s) \times (T \cdot S)	}$, where $N_t=S \cdot
(T-2)$ and $N_s=T \cdot (2n_rn_c-n_r)$ are the number of temporal and
spatial constraints, respectively. 
% The exact form of this matrix in clarified in Appendix
% B in the Supplement. 
Then, as we have two different tuning parameters
for the temporal and spatial components, we write $\Lambda
=\left[\lambda_t\one_{N_t}^\top,\;
  \lambda_s\one_{N_s}^\top\right]^\top$ leading
to:\footnote{Throughout the paper, we use $|x|$ for both scalars and
  vectors. For vectors we use this to denote a vector obtained by
  taking the absolute value of each entry of $x$.}  
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | Dh |.
\label{eq:l1tf_var_st_mat}
\end{equation}

@@

\autoref{fig:cities_ts} shows the time-series of the 
temperature of three cities: Indianapolis (USA), San Diego (USA) and
Manaus (Brazil). The time-series of Indianapolis and San Diego show
clear cyclic behavior, though the amplitude of the cycles
changes. The time-series of Manaus does not
show any regular cyclic behavior. For this reason, we
first apply trend filtering to remove seasonal terms and de-trend
every time series. For each time-series, we found the optimal value of
the penalty parameter using 5-fold cross-validation with
$k=5$. 
% We used the R package \textbf{genlasso} to perform these
% computations \cite{arnold_efficient_2016}.  
\begin{figure}[tb]
	\centering
	\includegraphics[width=.66\columnwidth]{Figures/cities_ts}
 	\caption{Time-series of the temperature (in Kelvin) of three cities.}
 	\label{fig:cities_ts}
\end{figure} 
The blue curve in the left panel of 
\autoref{fig:bloom_estimatedSD} shows the time-series of the
temperature of Indianapolis after detrending using this method.

\begin{figure}[tb]
  \centering
  \includegraphics[width=.8 \columnwidth]{Figures/indy_estimatedSD_1}
  \includegraphics[width=.45 \columnwidth]{Figures/indy_estimatedSD_2}
  \includegraphics[width=.45 \columnwidth]{Figures/indy_estimatedSD_3}  
  \caption{Top: The variability of the time-series of Indianapolis
    (weekly) and the estimated SD obtained from the method of
    \autoref{sec:l1tf_var} (red). Lower left: the estimated SDs (red) and
    their annual average (black) without imposing the long horizon
    penalty. Lower right: the same but with the long horizon
    penalty is imposed.} 
  \label{fig:bloom_estimatedSD}
\end{figure} 
The red curve 
shows the estimated SD (which is $\exp(h_t/2)$) obtained from out
proposed model. For ease of analysis, we compute the average
of the estimated SD for each year. Both are shown in the lower left panel of
\autoref{fig:bloom_estimatedSD}.

In addition to the constraints discussed in \autoref{sec:exten}, we
, we add
a long horizon penalty to smooth the annual trend.
$\sum_{i=1}^{N_{year}-2}\left| \sum_t h_{t_1}-2h_{t_2}+h_{t_3}\right|$
where $N_{year}$ is the number of years, $t_1=52(i-1)+t$, $t_2=52i+t$
and $t_3=52(i+1)+t$. Finally, because the observations are on the surface
of a hemisphere rather than a grid, we add extra spatial constraints with the
obvious form to handle the boundary between
$180^{\circ}$W and $180^{\circ}$E as well as the region near the North Pole.
The estimated SDs for Indianapolis are shown
the right panel of \autoref{fig:bloom_estimatedSD}. The annual average
of the estimated SDs shows a linear trend with a positive slope.

>>>>

\pdfinfo{
  /Title (Algorithms for Estimating Trends in Global Temperature Volatility)
   /Author (Khodadadi and McDonald)
  /Author (John Doe, Jane Doe)}

@@

\begin{abstract}
Trends in terrestrial temperature variability are perhaps more relevant for species
viability than trends in mean
temperature. In this paper, we develop methodology for estimating such
trends using multi-resolution climate data from polar orbiting weather
satellites. We derive two novel algorithms for computation that are
tailored for dense, gridded observations over both 
space and time. We evaluate our methods with a simulation constructed
to mimic these data's 
features and on a large, publicly available, global
temperature dataset with the eventual goal of tracking trends in
cloud reflectance temperature variability.
\end{abstract}

@@

Trends in terrestrial temperature variability are perhaps more relevant for species
viability than trends in mean
temperature~\cite{huntingford_no_2013},
because an increase in
temperature variability will increase the probability of extreme hot
outliers~\cite{VasseurDeLong2014}. Recent climate literature
suggests that it is more difficult for society to adapt to these
extremes than to the gradual increase in the mean temperature
\cite{hansen_perception_2012,huntingford_no_2013}. Furthermore, the willingness of popular media to
emphasize the prevalence extreme cold events coupled with a
fundamental misunderstanding of the relationship between climate (the global
distribution of weather over the long run) and weather (observed
short-term, localized behavior) leads to public misunderstanding of climate
change. In fact, a point of active debate is the extent to which the observed
increased frequency of extreme cold events in the 
northern hemisphere can be attributed to increases in temperature
variance  rather than to increases in
mean climate~\cite{Screen2014,FischerBeyerle2013,TrenberthZhang2014}.

@@

\citet{KimKoh2009} proposed $\ell_1$-trend filtering as a
method for estimating a smooth, time-varying trend. It is formulated
as the optimization problem 
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that
% it consists of a smooth trend and a stochastic component and the
% goal is to estimate the trend. The estimated trend is desired to be
% smooth, but at the same time, we seek small estimated residuals
% (stochastic component). These two objectives cannot be achieved
% simultaneously and so a trade-off between them should be made. In
% $\ell_1$-trend filtering, this trade-off is formulated as the
% following convex optimization problem: 
$$
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda
\sum_{t=1}^{T-2} \left|\beta_t-2\beta_{t+1}+\beta_{t+2}\right| 
$$
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \norm{ y-\beta }_2^2+\lambda \norm{ D_t \beta}_1
\label{eq:l1tf}
\end{equation}
 where $y=\{y_t\}_{t=1}^T$ is an observed time-series, $\beta\in \mathbb{R}^T$ is the smooth trend,
 $D_t$ is a $(T-2)\times T$ matrix (see Appendix C), and $\lambda$ is a tuning parameter
 which balances fidelity to the data (small errors in the first term)
 with a desire for smoothness.  
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
% In Appendix A, we specify the form of the matrix $D$ which results in piecewise
% linear estimated $\beta$. 
\citet{KimKoh2009} proposed a specialized primal-dual
interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a
statistical perspective, \eqref{eq:l1tf} can be viewed as a constrained maximum
likelihood problem with independent observations from a normal
distribution with common variance, $y_t \sim \mbox{N}(\beta_t,
\sigma^2)$, subject to a piecewise linear constraint on
$\beta$. Alternatively, solutions to \eqref{eq:l1tf} are %equivalent to
maximum a posteriori Bayesian estimators based on Gaussian likelihood
with a special Laplace prior distribution on $\beta$. Note that the
structure of the estimator is determined by the penalty function
$\lambda\norm{D_t\beta}_1$ rather than any parametric trend
assumptions---autoregressive, moving average, sinusoidal seasonal
component, etc. The resulting trend is therefore essentially
nonparametric in the same way that splines are nonparametric. In fact,
using squared $\ell_2$ norm as the penalty instead of $\ell_1$ results exactly in
regression splines.

@@

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend filtering on
graphs~\cite{WangSharpnack2016} which penalizes the difference between
the estimated values of the signal on the connected nodes (though the
likelihood is different). As before,
we can write \eqref{eq:l1tf_var_st} in matrix form where $h$ is a
vector of length
$TS$ and $D_t$ is replaced by $D \in
\mathbb{R}^{(N_t+N_s) \times (T \cdot S)	}$ (see Appendix C), where $N_t=S \cdot
(T-2)$ and $N_s=T \cdot (2n_rn_c-n_r)$ are the number of temporal and
spatial constraints, respectively. 
% The exact form of this matrix in clarified in Appendix
% B in the Supplement. 
Then, as we have two different tuning parameters
for the temporal and spatial components, we write $\Lambda
=\left[\lambda_t\one_{N_t}^\top,\;
  \lambda_s\one_{N_s}^\top\right]^\top$ leading
to:\footnote{Throughout the paper, we use $|x|$ for both scalars and
  vectors. For vectors we use this to denote a vector obtained by
  taking the absolute value of each entry of $x$.}  
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | Dh |.
\label{eq:l1tf_var_st_mat}
\end{equation}

@@

\autoref{fig:cities_ts} shows the time-series of the 
temperature of three cities: Indianapolis (USA), San Diego (USA) and
Manaus (Brazil). The time-series of Indianapolis and San Diego show
clear cyclic behavior, though the amplitude of the cycles
changes. The time-series of Manaus does not
show any regular cyclic behavior. For this reason, we
first apply trend filtering to remove seasonal terms and de-trend
every time series. For each time-series, we found the optimal value of
the penalty parameter using 5-fold cross-validation with
$k=5$. 
% We used the R package \textbf{genlasso} to perform these
% computations \cite{arnold_efficient_2016}.  
\begin{figure}[tb]
	\centering
	\includegraphics[width=.66\columnwidth]{Figures/cities_ts}
 	\caption{Time-series of the temperature (in Kelvin) of three cities.}
 	\label{fig:cities_ts}
\end{figure} 
The blue curve in the top panel of 
\autoref{fig:bloom_estimatedSD} shows the time-series of the
temperature of Indianapolis after detrending using this method.

\begin{figure}[tb]
  \centering
  \includegraphics[width=.8 \columnwidth]{Figures/indy_estimatedSD_1}
  \includegraphics[width=.45 \columnwidth]{Figures/indy_estimatedSD_2}
  \includegraphics[width=.45 \columnwidth]{Figures/indy_estimatedSD_3}  
  \caption{Top: The variability of the time-series of Indianapolis
    (weekly) and the estimated SD obtained from the method of
    \autoref{sec:l1tf_var} (red). Lower left: the estimated SDs (red) and
    their annual average (black) without imposing the long horizon
    penalty. Lower right: the same but with the long horizon
    penalty is imposed.} 
  \label{fig:bloom_estimatedSD}
\end{figure} 
The red curve 
shows the estimated SD (which is $\exp(h_t/2)$) obtained from our
proposed model. For ease of analysis, we compute the average
of the estimated SD for each year. Both are shown in the lower left panel of
\autoref{fig:bloom_estimatedSD}.

In addition to the constraints discussed in \autoref{sec:exten}, we
, we add
a long horizon penalty to smooth the annual trend.
$\sum_{i=1}^{N_{year}-2}\left| \sum_t h_{t_1}-2h_{t_2}+h_{t_3}\right|$
where $N_{year}$ is the number of years, $t_1=52(i-1)+t$, $t_2=52i+t$
and $t_3=52(i+1)+t$. Finally, because the observations are on the surface
of a hemisphere rather than a grid, we add extra spatial constraints with the
obvious form to handle the boundary between
$180^{\circ}$W and $180^{\circ}$E as well as the region near the North Pole.
The estimated SDs for Indianapolis are shown in the lower right panel of \autoref{fig:bloom_estimatedSD}. The annual average
of the estimated SDs shows a linear trend with a positive slope.
